---
title: "Data Analysis of IT jobs"
output: rmarkdown::github_document
---
Data Visualiization and Analysis from the data collected in the previous steps from jobs.ch and itjobs.ch. 
Some of the job listings are rated from a scale from 0,9 according to my interests and qualifications.

## Import Data
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
library(lubridate)
```
```{r}

raw_df <- fromJSON("../data/jobs_processed.json")
raw_df <- bind_rows(raw_df)
raw_df <- tibble(raw_df)
glimpse(raw_df)
```

## Clean Data
```{r, error=TRUE, warning=TRUE}
# convert datatypes
clean_df <- raw_df %>%
    mutate(publication_date = dmy(publication_date), 
        job_title_cleaned = factor(job_title_cleaned),
        career_stage_cleaned = factor(career_stage_cleaned),
        job_category = factor(job_category),
        canton = factor(canton),
        programming_languages = lapply(programming_languages, factor),
        frameworks = lapply(frameworks, factor),
        tools = lapply(tools, factor),
        operating_systems = lapply(operating_systems, factor),
        max_years = sapply(years, function(x) if (length(x) == 0) NA else max(as.numeric(x), na.rm = TRUE)), # get min number of years, and replace nan with 0
        programming_languages = ifelse(lengths(programming_languages) == 0, "No Programming Languages", programming_languages),
        frameworks = ifelse(lengths(frameworks) == 0, "No Frameworks", frameworks),
        tools = ifelse(lengths(tools) == 0, "No Tools", tools))
glimpse(clean_df)

```

Check if some IT key words in listing title were missed during preprocessing, and therefore have a NaN value in the `job_title_cleaned` attribute. 
```{r, error=TRUE, warning=TRUE}
clean_df %>%
  filter(is.na(job_title_cleaned)) %>%
  slice_head(n = 20) %>%
  pull(job_title) 
```
Keep only data with recognized IT- Keywords. Other job listings are probably not IT related.
```{r, error=TRUE, warning=TRUE}
# drop NaN rows
df <- clean_df %>%
    filter(!is.na(job_title_cleaned) & !is.na(publication_date))%>%
    filter(publication_date > as.Date("2025-01-01"))%>%
    filter(max_years <= 20) # outliers

    
glimpse(df)
```



## Visualize
Interesting to see the Sunndays with mostly zero job postings. Most postings are on Mondays and Thuesdays. 
```{r, time-jobs, error=TRUE, warning=TRUE}
df %>%
    count(publication_date) %>%
    ggplot(aes(publication_date, n)) +
    geom_line(aes(color = "All Listings"), linewidth = 1, alpha=0.5) +  # Line for all listings
    geom_line(data = df %>% filter(reviewed == TRUE) %>% count(publication_date), 
            aes(publication_date, n, color = "Reviewed Listings"), linewidth = 1, alpha=0.5) +  # Line for reviewed listings
    geom_point(aes(publication_date, n), alpha=0.5)+
    labs(title = "Number of scraped IT Job Listings per Day",
        x = "Date",
        y = "") +
    scale_color_manual(name = "Legend", values = c("All Listings" = "#1182ba", "Reviewed Listings" = "#71d171"))

```
```{r, error=TRUE, warning=TRUE}
df %>%
  ggplot() +
  geom_bar(mapping=aes(x=rating), alpha = 0.7)+
  labs(x = "Rating",
       y = "Count") 

```
Most Jobs are in the broad cattegory of Software Engineering, where some keywords of this category are: "Software Engineer", "Software Developer", "Full-Stack Developer", "Applikationsentwickler/in"...
Seccond most jobs are in the category of Cloud/System Engineering, where some keywords of this category are: "Cloud Engineer", "Cloud Architect", "Site Reliability Engineer", "System Engineer", "Systemadministrator"...

```{r, error=TRUE, warning=TRUE}
df %>%
  ggplot() +
  geom_bar(mapping=aes(y=job_category), alpha = 0.7)+
  labs(y = "Job Category",
       x = "Count") 

```
There are very few explicit junior positions. Senior positions have a mean of 3.9 years of experience, and positions with no explicit career stage have a mean of 3.1 years. Note there are some outliers.
```{r, error=TRUE, warning=TRUE}
df %>%
  group_by(career_stage_cleaned) %>%
  summarise(count = n(), mean_max_years = mean(max_years, na.rm = TRUE))
```
Junior positions, except for one outlier, always have zero years of required experience, which makes sense.
```{r, error=TRUE, warning=TRUE}
df %>% 
    ggplot(aes(x = max_years, y = career_stage_cleaned)) +
    geom_boxplot() +
    labs(y = "Career Stage",
         x = "Max years of experience",
         title = "Career Stage vs. Max Years of Experience with Mean Points")
```

```{r, error=TRUE, warning=TRUE}
df %>%
    unnest(programming_languages) %>%         # Unnest the list column
    count(programming_languages) %>%           # Count occurrences of each language
    ggplot(aes(x = n, y = programming_languages)) +  
    geom_bar(stat = "identity", alpha = 0.7) +
    labs(x = "Count", y = "Programming Language")

```
```{r, error=TRUE, warning=TRUE}
df %>%
    unnest(programming_languages) %>% 
    ggplot(aes(x = rating , y = programming_languages)) +
    geom_boxplot() +
    labs(x = "Rating", y = "Programming Languages")
```
```{r, error=TRUE, warning=TRUE}
df %>%
    count(search_query) 
```
```{r, error=TRUE, warning=TRUE}
df %>%
    count(job_title_cleaned) %>%
    arrange(desc(n))
```
```{r, error=TRUE, warning=TRUE}
df %>%
    count(job_category) %>%
    ggplot(aes(x = n, y = job_category)) +
    geom_bar(stat = "identity")

```
```{r, error=TRUE, warning=TRUE}
df %>%
    ggplot(aes(x = rating , y = job_category)) +
    geom_boxplot(alpha = 0.7) +
    labs(x = "Rating", y = "Job Category")
```
```{r, error=TRUE, warning=TRUE}
df %>%
    ggplot(aes(x = rating , y = canton)) +
    geom_boxplot(alpha = 0.7) +
    labs(x = "Rating", y = "Canton")
```

```{r, error=TRUE, warning=TRUE}
df %>%
    ggplot(aes(x = max_years, y = rating )) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(x = "Max Years of Experience", y = " Rating", title = "Correlation between Rating and Max Years of Experience")
```

## Fit Model

- Linear model is probably not the best for this usecase, classification with kNN / SVM might be more appropriate (TODO)
```{r, error=TRUE, warning=TRUE}

df_one_hot <- df %>%
    select(job_title, url, job_category, career_stage_cleaned, programming_languages, frameworks, max_years, rating ) %>% 
    # handle NA values as an "Non Value" factor
    mutate(
        job_category = factor(replace_na(as.character(job_category), "No Value")),
        career_stage_cleaned = factor(replace_na(as.character(career_stage_cleaned), "No Value")))
    
df_one_hot <- df_one_hot%>% 
    mutate(programming_languages = lapply(programming_languages, factor)) %>%
    unnest(programming_languages) %>%
    distinct() %>%
    mutate(temp = 1) %>%
    pivot_wider(names_from = programming_languages, values_from = temp, values_fill = list(temp = 0))


df_one_hot <- df_one_hot %>%   
    unnest(frameworks) %>%
    distinct() %>%
    mutate(temp = 1) %>%
    pivot_wider(names_from = frameworks, values_from = temp, values_fill = list(temp = 0))


glimpse(df_one_hot)

```
```{r, error=TRUE, warning=TRUE}
df_one_hot %>%
    mutate(has_rating = !is.na(rating)) %>%
    group_by(has_rating) %>%
    summarise(
        count = n())

```
```{r, error=TRUE, warning=TRUE}
# split train test set
df_model <- df_one_hot %>% filter(!is.na(rating))
df_to_predict <- df_one_hot %>% filter(is.na(rating))
set.seed(1)
sample <- sample(c(TRUE, FALSE), nrow(df_model), replace=TRUE, prob=c(0.8,0.2))
df_train  <- df_model[sample, ]
df_test   <- df_model[!sample, ]
```
### Linear Model
The linear model fits relatively well to the data (getting worse with more rated data -> overfitted ), whereas the categories Data Engineer and Junior Positions are positively correlated with the rating, as well as frontend frameworks like React.js and Next.js. 
On the other hand, the number of required years and the programming languages CSS are negatively correlated with the rating. 
Note that the large number of attributes compared to the relatively low number of data points might lead to overfitting. 
Especially some of the attributes have less than 5 data points, which makes this analysis not very valid. 
However, it is fun to see that it somewhat reflects my interests and capabilities.
```{r, error=TRUE, warning=TRUE}

# Fit a linear regression model
model <- lm(rating ~ . , data = df_train[,!colnames(df_train) %in% c("url", "job_title")])

# Summary of the model
summary(model)
```
```{r, error=TRUE, warning=TRUE}
# Predict on test set
predictions_test <- predict(model, newdata = df_test[,!colnames(df_test) %in% c("url", "job_title")])
df_test <- df_test %>% mutate(predicted_rating = predictions_test) 
df_test %>%
    ggplot(aes(x = rating, y = predicted_rating)) +
    geom_point(alpha=0.5, size = 3) +
    geom_abline(intercept = 0, slope = 1, color = "#2086b9", alpha = 0.5, linewidth = 2) +
    labs(x = "Actual Rating", y = "Predicted Rating")
```
```{r, error=TRUE, warning=TRUE}

# assert that there are no levels in the prediction data that are not in the model
#model$xlevels[["url"]] <- union(model$xlevels[["url"]], levels(df_one_hot$url))
model$xlevels[["job_category"]] <- union(model$xlevels[["job_category"]], levels(df_one_hot$job_category))
#model$xlevels[["career_stage_cleaned"]] <- union(model$xlevels[["career_stage_cleaned"]], levels(df_one_hot$career_stage_cleaned))
#model$xlevels[["programming_languages"]] <- union(model$xlevels[["programming_languages"]], levels(df_one_hot$programming_languages))
#model$xlevels[["frameworks"]] <- union(model$xlevels[["frameworks"]], levels(df_one_hot$frameworks))

# prediction from a rank-deficient fit may be misleading -> we need more data
predictions <- predict(model, newdata = df_to_predict[,!colnames(df_to_predict) %in% c("url", "job_title")])

# Add predictions to df_to_predict
df_to_predict <- df_to_predict %>%
    mutate(predicted_rating = predictions)

```

### Show most promising non-rated predictions
```{r, error=TRUE, warning=TRUE}
# Add a column with clickable URLs
df_to_predict <- df_to_predict %>%
    mutate(clickable_url = paste0("[", url, "](", url, ")"))

# Show most promising non-rated predictions with clickable URLs
df_to_predict %>%
    arrange(desc(predicted_rating)) %>%
    select(job_title, clickable_url, predicted_rating) %>%
    slice_head(n = 20) %>%
    knitr::kable()
```

### Show highest rated jobs
```{r, error=TRUE, warning=TRUE}
# Add a column with clickable URLs
df_model <- df_model %>%
    mutate(clickable_url = paste0("[", url, "](", url, ")"))

# Show most promising non-rated predictions with clickable URLs
df_model %>%
    arrange(desc(rating)) %>%
    select(job_title, clickable_url, rating) %>%
    slice_head(n = 20) %>%
    knitr::kable()
```